---
title: "Intro to Data Frames in R"
author: "Luana Lima"
date: "1/27/2021"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Data frame definition

A **data frame** is used for storing data tables. It a list a vector of equal length. 

When we import data to R, data frame is the preferred way for storing the data because columns can have different modes (character, numeric, integer, logical, complex).

## Data frame built-in example

Let's look into a built-in data frame from package "datasets" - cars. The data give the speed of cars and the distances taken to stop.

```{r cars}
cars
```

Note that it has 2 columns and 50 rows.

## Data frame columns

Suppose you want just the column speed. How would you access that data? 

```{r}
cars$speed
```

How would you store it on another object?

```{r}
car_speed <- cars$speed
car_speed
```


## Transforming object in a data frame

Is the new object you create a data frame?
```{r}
#Option 1
class(car_speed)

#Option 2
is.data.frame(car_speed)
```

How could you make it a data frame?

```{r set object as dataframe}
df_car_speed <- as.data.frame(car_speed)
df_car_speed
class(df_car_speed)
```

## Adding columns to a data frame

How could you add columns to *df_car_speed*?

One of the reasons we transform vectors in to data is so we can add columns.

```{r}

car_dist <- cars$dist

# We're adding the column distance (vector) to our dataframe car speed. This is done with the
# function bind
#Option 1

# The column bind is being assigned to a new object df
df <- cbind(df_car_speed,car_dist)  #similarly rows could be added using rbind()
class(df)
df

#Or Option 2 - transform into a data frame before binding
df_car_dist <- as.data.frame(car_dist)  #op2
df_opt2 <- cbind(df_car_speed,df_car_dist) #bind two dataframes
class(df_opt2)
df_opt2

```

Note that when we transformed the vector in a data frame the name of the vector became the column name.

```{r}
colnames(df)  #or simply names()
names(df)
```

## Creating a data frame

How would you create a data frame?

```{r}
#useful function data.frame()
#                       "name of the column" = data 
create_df <- data.frame("speed"=car_speed,"dist"=car_dist) #If you don't specify the name, car_speed will be te name of the column, etc. 
create_df
```


## Data frame functions

Some useful functions to use with data frames.

```{r}
ncol(df)  # number of columns of the dataframe 
nrow(df)  # number of rows
length(df)  #same as ncol
summary(df) #summary stats of the dataframe 
head(df) #show the first 6 rows of df

#If you know the number of the column you want you can refer to that to access column
df[,1]

#you could also use this notation to delete columns
df <- df[,-2]
df
```

