---
title: "Assignment 2: Coding Basics"
author: "Camila Zarate Ospina"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on coding basics. 

## Directions
1. Change "Student Name" on line 3 (above) with your name.
2. Work through the steps, **creating code and output** that fulfill each instruction.
3. Be sure to **answer the questions** in this assignment document.
4. When you have completed the assignment, **Knit** the text and code into a single PDF file.
5. After Knitting, submit the completed exercise (PDF file) to the dropbox in Sakai. Add your first and last name into the file name (e.g., "FirstLast_A02_CodingBasics.Rmd") prior to submission.

## Basics Day 1

1. Generate a sequence of numbers from one to 100, increasing by fours. Assign this sequence a name. 

```{r sequence}
# from, to, by
sequence <- seq(1, 100, 4) 
sequence
```

2. Compute the mean and median of this sequence. 

```{r}
meanseq <- mean(sequence)
medianseq <- median(sequence)

meanseq
medianseq
```

3. Ask R to determine whether the mean is greater than the median. 

```{r conditional}

meanseq > medianseq

```

4. Insert comments in your code to describe what you are doing.

```{r}
#1. Used the command seq to generate the sequence requested, following the "from, to, by" form. 

#2. Assigned the operation to obtain the mean and the median of the sequence to two objects: meanseq and medianseq, respectively.

#3. Used the command > and the objects created in the previous step to determine if one was greater than the other. 

```

## Basics Day 2
5. Create a series of vectors, each with four components, consisting of (a) names of students, (b) test scores out of a total 100 points, and (c) whether or not they have passed the test (TRUE or FALSE) with a passing grade of 50.

```{r create vectors}

names <- c("Amrita", "Natalia", "Megan", "Camila")
testscores <- c(40, 70, 80, 90)
pass <- c(FALSE, TRUE, TRUE, TRUE)

```

6. Label each vector with a comment on what type of vector it is.

```{r vector classes}

class(names) 
class(testscores)
class(pass)

```

7. Combine each of the vectors into a data frame. Assign the data frame an informative name.

```{r vectors into dataframe}

examresults.df<-data.frame(names, testscores, pass)

```

8. Label the columns of your data frame with informative titles.

```{r name dataframe columns and show dataframe}

names(examresults.df) <- c("Student name", "Test score", "Pass?")
View(examresults.df)

```

9. QUESTION: How is this data frame different from a matrix?

> Answer: A matrices contain elements of the same type, and a data frame can contain elements of different types, in our case, characters, numeric and logical.

10. Create a function with an if/else statement. Your function should determine whether a test score is a passing grade of 50 or above (TRUE or FALSE). You will need to choose either the `if` and `else` statements or the `ifelse` statement. Hint: Use `print`, not `return`. The name of your function should be informative.

11. Apply your function to the vector with test scores that you created in number 5. 

```{r}

score <- function(testscores){ifelse(testscores>49, TRUE, FALSE)}
score(100)

```

12. QUESTION: Which option of `if` and `else` vs. `ifelse` worked? Why?

> Answer: 
